---
title: "CCAMLR Decision Rules and Climate Change"
output: pdf_document
---

by Andrew Constable  
  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(deSolve)

source("CCAMLR Decision Rules - Functions.R")
source("CCAMLR Decision Rules - Input Data.R")

```

# 1. Introduction  
  
This routine provides a general Schaeffer production model to explore general principles associated with the CCAMLR Decision Rule for Antarctic krill fisheries.  
  
The decision rule has three parts, which say (SC-CAMLR 1989):  
  
(i) target level  
(ii) depletion level  
(iii) the lower of the catch limits will apply.  
  
The rule has three expectations:  
i) that sufficient krill escapes the fishery in order to provide food for predators, and that predators will be able to recovery from the effects of fishing within 2-3 decades.  
ii) recruitment of krill will be stable, such that fluctuations in the krill stock will not compromise krill recruitment in the long term, and that the krill stock will recover in 2-3 decades should the fishery be required to stop because it has reached critically low levels.  
  
This document is illustrative but intends to guide the reader through these principles as well as show when the decision rules may require adjustment because of climate change.  
  
# 2. Krill model

Krill is modelled as a biomass pool with production modelled as a Schaeffer production model.
In a food web, krill would be expected to have a biomass stable somewhere around its maximum production but greater than when maximum production occurs.  Evolutionary stable strategies of predators would expect to have long-term population responses to short-term prey shortages that ensure the predator population reduces its demand on krill to be less than maximum krill production and not over-consume krill to the point that the population would be in irreversible decline i.e. the predator population remains too high and consume too much krill relative to the krill population and its production at the those levels. Adjustments by predators could include eating less krill per capita, starving to death (reducing population consumption), or reducing reproductive output (recruitent rate).  The rate of reduction in consumption of krill would be fast, medium and slow adjustments respectively.   

The combined population effects of predator feeding, response to starvation, and adjustable reproductive rates on where the krill population will be stable can be explored using a simple predator-prey model that enables flexibility in productivity of krill as well as the different attributes of predators.

The key attributes of a balanced system are that the total predator consumption is a proportion of the maximum available production of krill, the krill population is able to recover from the size at maximum production to the size at balance with predators in the time of the expected maximum age of the krill population (i.e. the recruitment rate is still stable), and the intrinsic rate of increase of krill is sufficient to achieve both these requirements. 

```{r krill_model, echo=FALSE, eval=TRUE}
# 1.3 Calculate parameters from inputs ####
# (note: order is important)

EnvTimeSeries<-do.call(c,lapply(Env$Period,function(p){if(p$doFn) do.call(p$fn,list(p$years,p$data)) else rep(p$data,p$years)}))

Krill$Bmax_dP<-nlm(findBmax_dP,0.5*Krill$par$K,Krill$par)$estimate  # guess if for phi=1
Krill$B0<-optimise(findKrillB0,interval=c(Krill$Bmax_dP,Krill$par$K),Ppar=Krill$par ,bmax=Krill$Bmax_dP ,pmax=dP(Krill$Bmax_dP,Krill$par) ,p=Krill$B0asPropKrillMaxProd )$minimum

Krill$par$M<-dP(Krill$B0,Krill$par)/Krill$B0
Krill$B_pmax_givenM<-nlm(findProdMaxGivenM,Krill$Bmax_dP,Krill)$estimate

# 1.4 Illustrative plots ####
HollingExamples<-list( Whale   = Predators$Whale$Holling
                      ,Penguin = Predators$Penguin$Holling)
HollingExamples$Whale$p50<-HollingExamples$Whale$p50*Krill$B0  # correct scaling of p50 to new B0
HollingExamples$Penguin$p50<-HollingExamples$Penguin$p50*Krill$B0

p<-fnPlotPellaTomlinson(Krill,HollingExamples)
p
print(paste0("Krill M             = ",Krill$par$M))
print(paste0("Krill max Ingestion = ",Krill$par$Mmax))

```

# 3. Fishery

Currently catch limits are determined according to projections over a 10 year period to determine a constant catch that takes the population to target level over a 10-20 year projection.  (note here that fishing down a stock may be a higher rate using this application than might be implied for a 20 year period when the krill population is at the target level)


```{r fishery,echo=FALSE,eval=TRUE}
# 3.0 Illustrative plots based on assessment of gamma for population ####

# 3.1 Determine gamma for mortality rate

HoldUseMrate<-Krill$par$useMrate
Krill$par$useMrate<-TRUE
GammaM08<-optimise(find_G_Pop,c(0,1),Ppar=Krill,B0=Krill$B0,Btarget=Fishery$K_target_0*Krill$B0,projYrs=ProjYears$GammaCalcs)$minimum
Krill$par$useMrate<-FALSE
GammaHolling<-optimise(find_G_Pop,c(0,1),Ppar=Krill,B0=Krill$B0,Btarget=Fishery$K_target_0*Krill$B0,projYrs=ProjYears$GammaCalcs)$minimum
Krill$par$useMrate<-HoldUseMrate

print(paste0("Gamma using mortality rate = ",GammaM08))
print(paste0("Gamma using functional relationship = ",GammaHolling))

fnPlotPopWithGamma<-function(PT,ProjYears,GammaM=NULL,GammaHolling=NULL,MainPlotUsingGammaM=TRUE){

# project time series on each mortality type

plotYrs<-ProjYears$YearsTotal
parms<-list(PT=PT, catch=rep(0,ProjYears$YearsTotal))
y0<-PT$B0
names(y0)<-"Biomass"
pd<-NULL
if(!is.null(GammaM)){
parms$PT$par$useMrate<-TRUE
parms$catch[ProjYears$Fishery]<-GammaM*PT$B0
res<-as.data.frame(ode(y=y0,times=seq(0,plotYrs,0.2),func=fnPellaTomlinson_dBdt,parms=parms,method="rk4"))
Mortality<-rep("Rate",nrow(res))
pd<-rbind(pd,cbind(Mortality,res))
}
if(!is.null(GammaHolling)){
parms$PT$par$useMrate<-FALSE
parms$catch[ProjYears$Fishery]<-GammaHolling*PT$B0
res<-as.data.frame(ode(y=y0,times=seq(0,plotYrs,0.2),func=fnPellaTomlinson_dBdt,parms=parms,method="rk4"))
Mortality<-rep("Function",nrow(res))
pd<-rbind(pd,cbind(Mortality,res))
}

# standardise to status
pd$Biomass<-pd$Biomass/y0
Xlim<-c(0,ProjYears$YearsTotal)
#Ylim<-c(0,ceiling(max(pd$Biomass)*100)/100)
Ylim<-c(0,1)#max(pd$Biomass))

pd$Mortality<-factor(pd$Mortality,levels=unique(pd$Mortality))


p<-ggplot(pd)+labs(x="Year",y="Status")
p<-p+theme(panel.background = element_rect(fill="white",colour = "black")
           ,panel.grid.major = element_blank()
           ,panel.grid.minor = element_blank()
           ,axis.line=element_line(colour="black")
           ,axis.text = element_text(size=14)
           ,axis.title=element_text(size=18)
)# end theme
p<- p+scale_x_continuous(breaks=seq(Xlim[1],Xlim[2],length.out=6)
                         ,limits=Xlim
                         ,expand=expansion(mult = c(0,0.05)))
p<- p+scale_y_continuous(breaks=seq(Ylim[1],Ylim[2],length.out=5)
                         ,limits=Ylim
                         ,expand=expansion(mult = c(0,0.05)))
p<- p+guides(
  x = guide_axis(minor.ticks = TRUE)
  ,y = guide_axis(minor.ticks = TRUE)
)

p<- p+geom_hline(yintercept=c(0.2,0.75,1.0),colour=c("black","black","black"),linetype=c("dotted","dotted","dotted"))
p<- p+geom_vline(xintercept=c((ProjYears$Fishery[1]-1),ProjYears$Fishery[length(ProjYears$Fishery)],(ProjYears$Fishery[length(ProjYears$Fishery)]+21)),colour=c("black","black","black"),linetype=c("dashed","dashed","dashed"))

p<- p+geom_line(aes(x=time,y=Biomass,colour=Mortality))
return(p)
} # end fn
  
fnPlotPopWithGamma(Krill,ProjYears,GammaM=GammaM08,GammaHolling=GammaHolling)


```


#3. Predator model

To determine population size at carrying capacity, start predator population at a low level and let expand.  Compare to when krill has a variable carrying capacity and see what happens.

```{r pred_model,echo=FALSE,eval=TRUE}

# 2.0 setup predators ####
Pred<-lapply(Predators[UsePredators],function(pC,pP,B0multiplier){
                    pC$Holling$p50<- pP$B0*pC$Holling$p50 # scale Holling functional response to krill
                  # Predator B0: determined by krill production at Krill B0 and predator consumption/biomass
                    pC$B0<- B0multiplier*pC$BrelP
                    pC$QBmax<-nlm(findPredQBmax,(pC$QB*pC$B0),pC$QB,pC$B0,pP$B0,pC$Holling)$estimate
                    pC$M<-pC$PB
                    pC$RB<-pC$QB*pC$A-pC$PB
                    return(pC)
                   },Krill,(dP(Krill$B0,Krill$par)/sum(sapply(Predators[UsePredators],function(p){p$BrelP*p$QB}))))
names(Pred)<-UsePredators

# 2.1 Illustrative plots ####
PredColours<-as.vector(sapply(Pred,function(p){p$plot$col}))
xKrill<-seq(0,1,1E-3)*2*Krill$B0

res<-as.data.frame(do.call(rbind,lapply(Pred,function(p,Bp){
         Ingest<-fnHolling(Bp,p$B0,p$Holling)
         r1<-data.frame(rep(p$name,length(Ingest)),rep(Bp,length(Ingest)),Ingest)
         return(r1)
         },xKrill)))
names(res)<-c("Predator","Krill","Ingest")

res$Predator<-factor(res$Predator,levels = UsePredators)
p<-ggplot(res,aes(x=Krill,y=Ingest,colour=Predator))+geom_line()
p<-p+scale_color_manual(values=PredColours)
p<-p+geom_vline(xintercept = Krill$B0,col="black",linetype=5)
p


```


# 5. Trial projections

```{r PopProj,echo=FALSE,eval=TRUE}

# 4.1 Inputs ####

projYrs<-ProjYears$GammaCalcs


# 4.3 Calculate parameters from inputs ####
# 4.4 Illustrative plots ####

parms<-list(pE = EnvTimeSeries
           ,pP = Krill
           ,pC  = Pred
           ,catch = 0)

fwB0<-c(parms$pP$B0,as.vector(sapply(parms$pC,function(p){p$B0})))
names(fwB0)<-c(parms$pP$name,as.vector(sapply(parms$pC,function(p){p$name})))

HoldChangingK<-parms$pP$modelK$changingK
parms$pP$modelK$changingK <-FALSE
GammaFW<-optimise(find_G_FW,c(0,2*GammaM08),parms=parms,fwB0=fwB0,Btarget=Fishery$K_target_0*Krill$B0,projYrs=projYrs)$minimum

parms$catch<-rep(0,ProjYears$YearsTotal)
parms$catch[ProjYears$Fishery]<-GammaFW*Krill$B0

# parms$pP$modelK$changingK <-HoldChangingK

res<-ode(y=fwB0,times=seq(0,ProjYears$YearsTotal-1,0.2),func=fnFoodWeb_dBdt,parms=parms,method="rk4")


dPlot<-do.call(rbind,lapply(seq(2,ncol(res),1),function(s,B,B0){
         return(data.frame(Taxon=rep(dimnames(B)[[2]][s],nrow(B)),Year=B[,1],Status=B[,s]/B0[(s-1)]))
         },res,fwB0))
dPlot$Taxon<-factor(dPlot$Taxon,levels=unique(dPlot$Taxon))

Ylim<-c(0,max(dPlot$Status))
p<-ggplot(dPlot,aes(x=Year,y=Status,color=Taxon))+geom_line()
p<-p+scale_color_manual(values=as.vector(c(Krill$plot$col,sapply(Pred,function(p){p$plot$col}))))
p<-p+ylim(Ylim)+labs(title=paste0("Gamma FW = ",round(GammaFW,3)))
p




```

# 6. Scenarios

```{r Scenario_1, echo=FALSE, eval=TRUE}

```
