---
title: "SOweb - scenarios"
author: "Andrew Constable"
date: "2023-12-30"
output: pdf_document
bibliography: ["SOweb.bib"]
biblio-style: "ecology"
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(bookdown)
```

# Scenarios


## Results

```{r echo=FALSE,eval=FALSE}
n<-10 # number of species pools
B<-rep(1,N)
P<-rep(0.1,N) # per biomass production
M<-rep(0.001,N) # per biomass metabolism
W<-rep(0.0001,N) # per biomass waste of food - could be a matrix giving waste from each prey item
# a    matrix of availability of one taxon to another
# B.D=sum across all nodes[ aB(P+M+W) ]

# N+sum(bBW)=sum across all nodes[ aB(P+M+W) ]  # N = nutrients replenished in a year, b is the proportion of waste recycled
# D for terminal nodes are fixed


solveB<-function(B    # biomass vector to estimate
                ,N    # nutrients (if a vector)
                ,BT=0 # terminal node biomass if needed
                ,p    # per bimoass productivity
                ,m    # per biomass metabolism
                ,w    # per biomass waste
                ,a    # matrix of availabilities (Ecotrophic Efficiencies in Ecopath)
                ,b    # vector of waste recycling
){
  # loop through Nodes not including nutrients
  # calculate N =  sum across all nodes[ aB(P+M+W) ] - sum(bBW) 
}
```

Include competition

Randomise over uncertainty including starting vector
Do PCA on results to show main forces and orientations.  Need a centroid to display the food web structure
In the function, can include part of starting vector as absolute or relative abundances with error
Can you do a likelihood of a vector based on what the linkages would need to be to make it work (departure of linkages from estimated linkages))




