---
title: "CCAMLR Decision Rules and Climate Change"
output: pdf_document
---

by Andrew Constable  
  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggpubr)
library(deSolve)

source("CCAMLR Decision Rules - Functions.R")
source("CCAMLR Decision Rules - Input Data.R")

```

# Notes for Assessing CCAMLR decision rule
Introduction
A. current rule & expectations
B. predator requirements
C. dealing with growth and uncertainty - develop relative importance of each
D. envirinmental variability and climate change
E. current approach
i. swarms as a form of hyperstability in krill density
ii. local escapement versus global reduction
F. what happens over different lengths of time
i. recovery of predators 
ii. is the escapement rule appropriate for which median


Methods
A. model to support assessment
a. modelling environment to be able to insert climate modelling
i. variable phytoplankton
ii. vary the availability of the pool to krill cf copepods

b. modelling krill and copepods
i. surplus pdroduction
ii. biomass with density dependent recruitment (2004 model)
iii. age structured

c. modelling predators
i.  Ecopath-like model - Predator-production with instantaneous mortality due to starvation
ii. Predator-production with delayed mortality due to starvation having gradual effect on mass
iii. Predator-production as individual model with density-dependent recruitment 

B. management system
i. krill rule application
ii. predator requirements & rule application


Results
i. the current rule and its expectations
ii. variability and growth - which median to use once at target level (at end vs over whole projection period)
iii. predator requirements and 75%
iv. recovery of whales and climate change
v. a new rule for predators
vi. feedback metrics and adjustments (birth rate without density-dependent adjustment, population, pop-adj-diet)




# 1. Introduction  
  
This routine provides a general Schaeffer production model to explore general principles associated with the CCAMLR Decision Rule for Antarctic krill fisheries.  
  
The decision rule has three parts, which say (SC-CAMLR 1989):  
  
(i) target level  
(ii) depletion level  
(iii) the lower of the catch limits will apply.  
  
The rule has three expectations:  
i) that sufficient krill escapes the fishery in order to provide food for predators, and that predators will be able to recovery from the effects of fishing within 2-3 decades.  
ii) recruitment of krill will be stable, such that fluctuations in the krill stock will not compromise krill recruitment in the long term, and that the krill stock will recover in 2-3 decades should the fishery be required to stop because it has reached critically low levels.  
  
This document is illustrative but intends to guide the reader through these principles as well as show when the decision rules may require adjustment because of climate change.  
  
# 2. Krill model

Krill is modelled as a biomass pool with production modelled as a Schaeffer production model.
In a food web, krill would be expected to have a biomass stable somewhere around its maximum production but greater than when maximum production occurs.  Evolutionary stable strategies of predators would expect to have long-term population responses to short-term prey shortages that ensure the predator population reduces its demand on krill to be less than maximum krill production and not over-consume krill to the point that the population would be in irreversible decline i.e. the predator population remains too high and consume too much krill relative to the krill population and its production at the those levels. Adjustments by predators could include eating less krill per capita, starving to death (reducing population consumption), or reducing reproductive output (recruitent rate).  The rate of reduction in consumption of krill would be fast, medium and slow adjustments respectively.   

The combined population effects of predator feeding, response to starvation, and adjustable reproductive rates on where the krill population will be stable can be explored using a simple predator-prey model that enables flexibility in productivity of krill as well as the different attributes of predators.

The key attributes of a balanced system are that the total predator consumption is a proportion of the maximum available production of krill, the krill population is able to recover from the size at maximum production to the size at balance with predators in the time of the expected maximum age of the krill population (i.e. the recruitment rate is still stable), and the intrinsic rate of increase of krill is sufficient to achieve both these requirements. 


Figure 9.7b shows a typical fishery production model and the expectation of how much production taken from the krill stock by predators with a constant mortality rate.  Mortality is expected to equal production at the equilibrium biomass of the Ecopath model, equivalent to the pre-exploitation biomass referred to as B0.  

Figure 9.7 illustrates a number of issues that are problematic in the current assessment. In each scenario, the system is projected through a period with no fishing, followed by a fishing period for 20 years under a constant catch using gamma from the assessment with constant mortality rate for krill (the current assessment approach), and a recovery period for 30 years. Two predators are represented – a long-lived species (whales) and a short-lived one, penguins. The parameters (annual per biomass consumption, productivity, mortality) for these predators and for krill were derived from the standardised Ecopath models, the best science available on food web parameters (Suppl. Material). The functional relationships of the predators were approximated from the ecologies of these species (Bestley et al., 2020).

```{r krill_model, echo=FALSE, eval=TRUE}
# 1.3 Calculate parameters from inputs ####
# (note: order is important)

plotColours<-as.vector(c(Env$plot$col,Krill$plot$col,sapply(Predators,function(p){p$plot$col}),Fishery$plot$col))
names(plotColours)<-c(Env$name,Krill$name,sapply(Predators,function(p){p$name}),Fishery$name)
plotLineType<-as.vector(c(Env$plot$linetype,Krill$plot$linetype,sapply(Predators,function(p){p$plot$linetype}),Fishery$plot$linetype))
names(plotLineType)<-c(Env$name,Krill$name,sapply(Predators,function(p){p$name}),Fishery$name)

EnvTimeSeries<-do.call(c,lapply(Env$Period,function(p){if(p$doFn) do.call(p$fn,list(p$years,p$data)) else rep(p$data,p$years)}))


Krill$Bmax_dP<-nlm(findBmax_dP,0.5*Krill$par$K,Krill$par)$estimate  # guess if for phi=1
Krill$B0<-if(Krill$useMtoFindB0){
    optimise(findKrillB0_useM,interval=c(Krill$Bmax_dP,Krill$par$K)
             ,Ppar=Krill$par)$minimum
   } else {
  optimise(findKrillB0_propPmax,interval=c(Krill$Bmax_dP,Krill$par$K)
           ,Ppar=Krill$par ,pmax=dP(Krill$Bmax_dP,Krill$par) 
           ,p=Krill$B0asPropKrillMaxProd )$minimum
   } # end findB0


Krill$par$M<-dP(Krill$B0,Krill$par)/Krill$B0
if(!Krill$par$useMrate) Krill$par$Mmax<-nlm(findMmax,dP(Krill$B0,Krill$par),Krill,Krill$B0,dP(Krill$B0,Krill$par))$estimate

Krill$B_pmax_givenM<-nlm(findProdMaxGivenM,Krill$Bmax_dP,Krill)$estimate

# 1.4 Illustrative plots ####
HollingExamples<-list( Whale   = Predators$Whale$Holling
                      ,Penguin = Predators$Penguin$Holling)
HollingExamples$Whale$p50<-HollingExamples$Whale$p50*Krill$B0  # correct scaling of p50 to new B0
HollingExamples$Penguin$p50<-HollingExamples$Penguin$p50*Krill$B0

p1<-fnPlotPellaTomlinson(Krill,plotColours,plotLineType,HollingExamples,Ylim=c(0,80))
p1
print(paste0("Krill M             = ",Krill$par$M))
print(paste0("Krill max Mortality by mass = ",Krill$par$Mmax))



```

# 3. Fishery

Currently catch limits are determined according to projections over a 10 year period to determine a constant catch that takes the population to target level over a 10-20 year projection.  (note here that fishing down a stock may be a higher rate using this application than might be implied for a 20 year period when the krill population is at the target level)


```{r fishery,echo=FALSE,eval=TRUE}
# determine gamma

whichGamma<-c("Rate","Whale","Penguin")

Gamma<-do.call(c,lapply(whichGamma,function(whichG,H,PT,Target,Yrs){
    if(whichG=="Rate"){
        PT$par$useMrate<-TRUE
        return(optimise(find_G_Pop,c(0,1)
                        ,PT=PT,B0=PT$B0,Btarget=Target*PT$B0
                        ,projYrs=Yrs)$minimum)
    } else {
                   PT$par$Holling<-H[[whichG]]
                   PT$par$useMrate<-FALSE
                   PT$par$Mmax<-nlm(findMmax,dP(PT$B0,PT$par),PT,PT$B0,dP(PT$B0,PT$par))$estimate
                   return(optimise(find_G_Pop,c(0,0.5)
                        ,PT=PT,B0=PT$B0,Btarget=Target*PT$B0
                        ,projYrs=Yrs)$minimum)
    } # end else
    },HollingExamples,Krill,Fishery$K_target_0,ProjYears$GammaCalcs))           
names(Gamma)<-whichGamma



parms<-list(PT=Krill, Btarget=Fishery$K_target_0*Krill$B0 , HollingExamples= HollingExamples, catch=NULL, ProjYears = ProjYears)

p2<-fnPlotPopWithGamma(Gamma,parms=parms,ProjYears=ProjYears,time0=-20,Xlim=c(0,ProjYears$YearsTotal),XticksN=8,Ylim=c(0,1.25),YticksN = 6,LegendTopLeft=c(0.83,0.165))

suppressWarnings(print(p2))

```


#3. Predator model

To determine population size at carrying capacity, start predator population at a low level and let expand.  Compare to when krill has a variable carrying capacity and see what happens.

```{r pred_model,echo=FALSE,eval=TRUE}

# 2.0 setup predators ####
Pred<-lapply(Predators,function(pC,pP){
                    pC$Holling$p50<- pP$B0*pC$Holling$p50 # scale Holling functional response to krill
                  # Predator B0: determined by krill production at Krill B0 and predator consumption/biomass
                    pC$B0<- NA
                    pC$QBmax<-NA
                    pC$M<-pC$PB
                    pC$RB<-pC$QB*pC$A-pC$PB
                    return(pC)
                   },Krill)
names(Pred)<-names(Predators)

# 2.1 Illustrative plots ####
PredColours<-as.vector(sapply(Pred,function(p){p$plot$col}))
xKrill<-seq(0,1,1E-3)*2*Krill$B0

res<-as.data.frame(do.call(rbind,lapply(Pred,function(p,Bp){
         Ingest<-fnHolling(Bp,p$B0,p$Holling)
         r1<-data.frame(rep(p$name,length(Ingest)),rep(Bp,length(Ingest)),Ingest)
         return(r1)
         },xKrill)))
names(res)<-c("Predator","Krill","Ingest")

res$Predator<-factor(res$Predator,levels = unique(res$Predator))
p<-ggplot(res,aes(x=Krill,y=Ingest,colour=Predator))+geom_line()
p<-p+scale_color_manual(values=PredColours)
p<-p+geom_vline(xintercept = Krill$B0,col="black",linetype=5)
p


```


# 5. Scenarios

## 5.1. Single predator-krill examples

Determine each gamma for the krill fishery based on the respective predator attributes. Here the decision rule is applied such that annual catch as a proportion of Krill B0 is the minimum catch to for the krill population to be reduced to the target level.


```{r sSinglePredator_Krill,echo=FALSE,eval=TRUE}

parms<-list(pE = EnvTimeSeries
           ,pP = Krill
           ,pC  = Pred
           ,pF  = Fishery
           ,catch = 0)

parms$pP$modelK$changingK <-FALSE  # environment remains constant for the duration

# update each predator B0 and QBmax as if only predator in food web
parms$pC<-lapply(parms$pC,function(pred,PreyB0,PreyProd){
  pred$B0<-PreyProd/pred$QB
  pred$QBmax<-nlm(findPredQBmax,(pred$QB*pred$B0),pred$QB,pred$B0,PreyB0,pred$Holling)$estimate
  return(pred)},parms$pP$B0,dP(parms$pP$B0,parms$pP$par))
names(parms$pC)<-names(Pred)

estGfw<-FALSE
if(estGfw){
   GammaFWsinglePred<-as.vector(sapply(Pred,function(p,parms,Btarget,projYrs){
       parms$pC<-list(p)
       fwB0<-c(parms$pP$B0,as.vector(sapply(parms$pC,function(p){p$B0})))  # biomasses of krill/preds time 0 (from Ecopath)
       names(fwB0)<-c(parms$pP$name,as.vector(sapply(parms$pC,function(p){p$name})))
       return(optimise(find_G_FW,c(0,0.5),parms=parms,fwB0=fwB0,Btarget=Btarget,projYrs=projYrs)$minimum)
       },parms,Fishery$K_target_0*Krill$B0,ProjYears$GammaCalcs))
    names(GammaFWsinglePred)<-names(Pred)
} else {
  GammaFWsinglePred<-Gamma[c("Whale","Penguin")]
}

GammaFW<-rep(Gamma[1],2) 

p3<-fnPlotFWwithGammaSinglePreds(GammaFW
,parms,ProjYears,time0=-20,Xlim=c(0,ProjYears$YearsTotal),XticksN=8,Ylim=c(0,1.25)
,YticksN=6,LegendTopLeft=c(0.82,0.2),doAxisLabels=TRUE,LineColours=plotColours)

suppressWarnings(print(p3))



```

## 5.2. Two predator-krill examples

```{r sTwoPredator_Krill,echo=FALSE,eval=TRUE}
parms<-list(pE = EnvTimeSeries
           ,pP = Krill
           ,pC  = Pred
           ,catch = 0)

parms$pP$modelK$changingK <-FALSE  # environment remains constant for the duration

# update each predator B0 assuming all predators in food web at proportions relative to krill abundance
# note that this procedure corrects for any relative problems from Ecopath model

B0multiplier<-(dP(Krill$B0,Krill$par)/sum(sapply(Predators,function(p){p$BrelP*p$QB})))
parms$pC<-lapply(parms$pC,function(pred,B0multiplier,PreyB0){
  pred$B0<-pred$BrelP*B0multiplier
  pred$QBmax<-nlm(findPredQBmax,(pred$QB*pred$B0),pred$QB,pred$B0,PreyB0,pred$Holling)$estimate
  return(pred)},B0multiplier,parms$pP$B0)
names(parms$pC)<-names(Pred)

estGfw<-FALSE
if(estGfw){
       fwB0<-c(parms$pP$B0,as.vector(sapply(parms$pC,function(p){p$B0})))  # biomasses of krill/preds time 0 (from Ecopath)
       names(fwB0)<-c(parms$pP$name,as.vector(sapply(parms$pC,function(p){p$name})))
       GammaFW<-optimise(find_G_FW,c(0,1),parms=parms,fwB0=fwB0,Btarget=Fishery$K_target_0*Krill$B0
                       ,projYrs=ProjYears$GammaCalcs)$minimum
} # end if

# plot productivity for enviornment at beginning and end and mortality functions at beginning
p<-fnPlotPellaTomlinson_FWconfig(c(100,80),parms$pP,parms$pC)
p


# 4 Foodweb, fishing, no change environment, whales at B0
GammaFW<-Gamma[1]
ChangingK<-FALSE
StartWhalesRelToB0<-1

p4<-fnPlotFWwithGammaTwoPreds(GammaFW
,parms,ProjYears,ChangingK=ChangingK,startWhalesX=StartWhalesRelToB0,Xlim=c(0,ProjYears$YearsTotal),XticksN=7,Ylim=c(0,1.25)
,YticksN=6,LegendTopLeft=c(0.85,0.18),doAxisLabels=TRUE,LineColours=plotColours)

suppressWarnings(print(p4))


# 5 Foodweb, no fishing, changing environment, whales at B0

GammaFW<-0
ChangingK<-TRUE
StartWhalesRelToB0<-1

p5<-fnPlotFWwithGammaTwoPreds(GammaFW
,parms,ProjYears,ChangingK=ChangingK,startWhalesX=StartWhalesRelToB0,Xlim=c(0,ProjYears$YearsTotal),XticksN=7,Ylim=c(0,3)
,YticksN=6,useStatus=TRUE,LegendTopLeft=c(0.87,0.18),doAxisLabels=TRUE,LineColours=plotColours)

suppressWarnings(print(p5))

# 6 Foodweb, fishing, changing environment, whales at B0

GammaFW<-Gamma[1]
ChangingK<-TRUE
StartWhalesRelToB0<-1

p6<-fnPlotFWwithGammaTwoPreds(GammaFW
,parms,ProjYears,ChangingK=ChangingK,startWhalesX=StartWhalesRelToB0,Xlim=c(0,ProjYears$YearsTotal),XticksN=7,Ylim=c(0,1.25)
,YticksN=6,LegendTopLeft=c(0.87,0.18),doAxisLabels=TRUE,LineColours=plotColours)

suppressWarnings(print(p6))

# 7 Foodweb, no fishing, no change environment, whales at 0.5*B0

GammaFW<-0
ChangingK<-FALSE
StartWhalesRelToB0<-0.5

p7<-fnPlotFWwithGammaTwoPreds(GammaFW
,parms,ProjYears,ChangingK=ChangingK,startWhalesX=StartWhalesRelToB0,Xlim=c(0,ProjYears$YearsTotal),XticksN=7,Ylim=c(0,2.0)
,YticksN=5,LegendTopLeft=c(0.85,0.18),doAxisLabels=TRUE,LineColours=plotColours)

suppressWarnings(print(p7))

# 8 Foodweb, fishing, no change environment, whales at 0.5*B0

GammaFW<-Gamma[1]
ChangingK<-FALSE
StartWhalesRelToB0<-0.5

p8<-fnPlotFWwithGammaTwoPreds(GammaFW
,parms,ProjYears,ChangingK=ChangingK,startWhalesX=StartWhalesRelToB0,Xlim=c(0,ProjYears$YearsTotal),XticksN=7,Ylim=c(0,2.0)
,YticksN=5,LegendTopLeft=c(0.85,0.18),doAxisLabels=TRUE,LineColours=plotColours)

suppressWarnings(print(p8))

# 9 Foodweb, no fishing, changing environment, whales at 0.5 B0

GammaFW<-0
ChangingK<-TRUE
StartWhalesRelToB0<-0.5

p9<-fnPlotFWwithGammaTwoPreds(GammaFW
,parms,ProjYears,ChangingK=ChangingK,startWhalesX=StartWhalesRelToB0,Xlim=c(0,ProjYears$YearsTotal),XticksN=7,Ylim=c(0,2.0)
,YticksN=5,LegendTopLeft=c(0.87,0.18),doAxisLabels=TRUE,LineColours=plotColours)

suppressWarnings(print(p9))

# 10 Foodweb, fishing, changing environment, whales at 0.5 B0

GammaFW<-Gamma[1]
ChangingK<-TRUE
StartWhalesRelToB0<-0.5

p10<-fnPlotFWwithGammaTwoPreds(GammaFW
,parms,ProjYears,ChangingK=ChangingK,startWhalesX=StartWhalesRelToB0,Xlim=c(0,ProjYears$YearsTotal),XticksN=7,Ylim=c(0,2.0)
,YticksN=5,LegendTopLeft=c(0.87,0.18),doAxisLabels=TRUE,LineColours=plotColours)

suppressWarnings(print(p10))

```


```{r graph_outputs,echo=FALSE,eval=TRUE}
pWidth<-7
pHeight<-5
pDPI<-300
pList<-list(p1  = list(p1 ,"Ecopath 01 Production")
           ,p2  = list(p2 ,"Ecopath 02 Gamma")
           ,p3  = list(p3 ,"Ecopath 03 FW 1 spp fishery")
           ,p4  = list(p4 ,"Ecopath 04 FW 2 spp fishery")
           ,p5  = list(p5 ,"Ecopath 05 FW no fishery change")
           ,p6  = list(p6 ,"Ecopath 06 FW fishery change")
           ,p7  = list(p7 ,"Ecopath 07 FW no fishery no change half whale")
           ,p8  = list(p8 ,"Ecopath 08 FW fishery no change half whale")
           ,p9  = list(p9 ,"Ecopath 09 FW no fishery change half whale")
           ,p10 = list(p10,"Ecopath 10 FW fishery change half whale"))
lapply(pList,function(p,W,H,D){
  ggsave(filename=paste0(p[[2]],".png"),plot = p[[1]]
      ,width = W,height = H,units = "cm",dpi = D)
      },pWidth,pHeight,pDPI)

```

```{r pagePlot, echo=FALSE, eval=TRUE}
ggarrange( NULL
          ,NULL
          ,p2 + rremove("xlab") + rremove("ylab") + rremove("x.text")  + font("y.text", size = 10) + rremove("legend")
          ,p1 + font("x.text", size = 10)  + font("y.text", size = 10)  + font("y.title", size = 10)  + font("x.title", size = 10) + rremove("legend")
          ,NULL
          ,NULL
          ,p3 + rremove("xlab") + rremove("ylab") + rremove("x.text")  + font("y.text", size = 10) + rremove("legend")
          ,p4 + rremove("xlab") + rremove("ylab") + rremove("x.text")  + font("y.text", size = 10) + rremove("legend")

          ,NULL
          ,NULL
          ,p7 + rremove("xlab") + rremove("ylab") + rremove("x.text")  + font("y.text", size = 10) + rremove("legend")
          ,p8 + rremove("xlab") + rremove("ylab") + rremove("x.text")  + font("y.text", size = 10) + rremove("legend")

          ,NULL
          ,NULL
          ,p5 + rremove("xlab") + rremove("ylab") + rremove("x.text")  + font("y.text", size = 10) + rremove("legend")
          ,p6 + rremove("xlab") + rremove("ylab") + rremove("x.text")  + font("y.text", size = 10) + rremove("legend")

          ,NULL
          ,NULL
          ,p9 + rremove("ylab")  + font("x.text", size = 10)  + font("y.text", size = 10) + rremove("legend")
          ,p10 + rremove("ylab") + font("x.text", size = 10)  + font("y.text", size = 10) + rremove("legend")
          ,labels = c("a. Assessment projection", "b. Krill productivity & mortality"
                      , "", ""
                      , "c. Krill - single predator scenarios", "d. Krill - two predator scenario"
                      , "", ""
                      , "e. Whale recovery G=0", "f. Whale recovery, G=..."
                      , "", ""
                      , "g. Production change, G=0", "h. Production change, G=..."
                      , "", ""
                      , "i. Production change, Whales, G=0", "j. Production change, Whales, G=...")
                      , "", ""
          , align = "v"
          ,heights=c(0.1,1,0.1,1,0.1,1,0.1,1,0.1,1.5)
          ,ncol = 2, nrow = 10)

```

